Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 paper.tex
Backwards in time

In this paper we provide a new formulation of ARG likelihood which
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs. \revpoint{2}{13}
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes
in the graph is not necessary
(from the perspective of computing the likelihood)
as long as the ``locally
unary''~\citep{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained.
We also show via an empirical example that the likelihood is
robust to the presence of polytomies and other imprecise
information, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

% \section{Methods and Results}
% First,
% we describe
% in more detail what sort of information we have in estimated ARGs.
% Then,
% we define the SMC process in a ``backwards-in-time'' way,
% rather than the usual ``along-the-genome'' way.
% Note that this is a
% reiteration of the original work of \citet{mcvean_approximating_2005}.

% In the subsequent section we connect the two and
% show how the ``backwards-in-time'' description
% makes it relatively straightforward to compute the likelihood
% for a more general class of estimated ARGs.
% Finally, we close with a demonstration that the likelihood computation can be used
% to do Bayesian inference on ancestral population sizes.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{SMC backwards-in-time}\label{par:description}

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
%
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm
to simulate the coalescent with
recombination was formulated backwards in time \revpoint{2}{24}
by Hudson \citeyearpar{hudson_properties_1983}.
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process
\citep{wiuf_recombination_1999}.
This along-the-genome approach is considerably more complex as the next
tree to be simulated depends on all previous trees.
This idea however provided the basis of the
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time,
as well as along-the-genome. \revpoint{2}{25}
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping}
genomic intervals may coalesce. The SMC' is a refinement
of the SMC which also allows lineages with abutting genomic intervals \revpoint{2}{26}
to coalesce~\citep{marjoram_fast_2006}.
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.

@@

% The second and probably the most important application of the ability to associate a more
% general class of ARGs to a principled model is the ability to boost the current efforts
% in ARG inference.
% % great starting point for MCMC
% Any inferred ARG can be used a good starting point
% for a future MCMC-sampler. This potentially implies a massive runtime reduction by
% reducing the necessary burn-in period. Furthermore, by integrating out the exact time
% to the recombination event we further restrict the size of the ARG space that needs to
% be explored. This MCMC-sampler would again rely on the tree sequence data structure to
% efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
% % hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
% To further improve scalability, hybrid ARG-inference methods could be designed which
% rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
% and subsequently use a principled approach to perform ARG inference for the second phase.
% This idea resembles the tried and tested approach of hybrid backwards-in-time simulations
% that combine Wright-Fisher dynamics in the very recent past and coalescent simulations
% for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
% approach has enabled large scale simulations while avoiding the documented biases of the
% coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}.
% The same approach could be thus used for ARG-inference where one can rely on a heuristic
% inference method for the recent past, while using a model-based approach beyond that cutoff point.

@@

% this implies that the backwards-in-time description of the

>>>>

In this paper we provide a new formulation of ARG likelihood which
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs. \revpoint{2}{13}
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards in time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes
in the graph is not necessary
(from the perspective of computing the likelihood)
as long as the ``locally
unary''~\citep{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained.
We also show via an empirical example that the likelihood is
robust to the presence of polytomies and other imprecise
information, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

% \section{Methods and Results}
% First,
% we describe
% in more detail what sort of information we have in estimated ARGs.
% Then,
% we define the SMC process in a ``backwards in time'' way,
% rather than the usual ``along-the-genome'' way.
% Note that this is a
% reiteration of the original work of \citet{mcvean_approximating_2005}.

% In the subsequent section we connect the two and
% show how the ``backwards in time'' description
% makes it relatively straightforward to compute the likelihood
% for a more general class of estimated ARGs.
% Finally, we close with a demonstration that the likelihood computation can be used
% to do Bayesian inference on ancestral population sizes.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{SMC backwards in time}\label{par:description}

% coalescent introduction: backwards in time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
%
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm
to simulate the coalescent with
recombination was formulated backwards in time \revpoint{2}{24}
by Hudson \citeyearpar{hudson_properties_1983}.
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process
\citep{wiuf_recombination_1999}.
This along-the-genome approach is considerably more complex as the next
tree to be simulated depends on all previous trees.
This idea however provided the basis of the
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards in time,
as well as along-the-genome. \revpoint{2}{25}
In its backwards in time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping}
genomic intervals may coalesce. The SMC' is a refinement
of the SMC which also allows lineages with abutting genomic intervals \revpoint{2}{26}
to coalesce~\citep{marjoram_fast_2006}.
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards in time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.

@@

% The second and probably the most important application of the ability to associate a more
% general class of ARGs to a principled model is the ability to boost the current efforts
% in ARG inference.
% % great starting point for MCMC
% Any inferred ARG can be used a good starting point
% for a future MCMC-sampler. This potentially implies a massive runtime reduction by
% reducing the necessary burn-in period. Furthermore, by integrating out the exact time
% to the recombination event we further restrict the size of the ARG space that needs to
% be explored. This MCMC-sampler would again rely on the tree sequence data structure to
% efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
% % hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
% To further improve scalability, hybrid ARG-inference methods could be designed which
% rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
% and subsequently use a principled approach to perform ARG inference for the second phase.
% This idea resembles the tried and tested approach of hybrid backwards in time simulations
% that combine Wright-Fisher dynamics in the very recent past and coalescent simulations
% for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
% approach has enabled large scale simulations while avoiding the documented biases of the
% coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}.
% The same approach could be thus used for ARG-inference where one can rely on a heuristic
% inference method for the recent past, while using a model-based approach beyond that cutoff point.

@@

% this implies that the backwards in time description of the
